Reasoning About a Capability Machine with Local Capabilities

### 2017-10-10 esop-2018/paper.tex
sec. 4

\section{Logical relation}
\label{sec:logical-relation}
Now that we have defined our calling convention, how can we be sure that it
works? More concretely, suppose that we have a program that uses the convention
in its interaction with untrusted adversary code. Can we formally prove the
program's correctness if it relies on well-bracketed control flow and private
state encapsulation for the interaction with the adversary? Clearly, such a
proof should depend on a formal expression of the guarantees provided by the
capability machine, including the specific guarantees for local capabilities.

In this section, we construct such a formalization. We make use of some
well-studied and powerful (but non-trivial) machinery from the literature.
Specifically, we employ a unary step-indexed Kripke logical relation with
recursive worlds, and some additional special characteristics of our own.
Step-indexing, Kripke logical relations and recursive worlds are techniques that
may be familiar from lambda calculus settings, but it may not be clear to the
reader how they apply in this more low-level assembly language. Therefore, we do
not immediately dive into the details, but we first try to provide some informal
intuition about how all of this machinery comes into play in our setting, in the
next section.

Note: even though the calling convention is the main application in this paper,
the logical relation we construct is very general and should be regarded as an
independent contribution.

\subsection{Formalizing the guarantees of the capability machine}
\label{sec:formalizing-guarantees}
What differentiates a capability machine from a more standard assembly language
is that we can bound the authority of an executing block of code, based solely
on the capabilities it has access to. Specifically, it does not matter which
instructions are actually executed, i.e., the bound also applies to untrusted
adversary code that has not been inspected or modified in any way.

\paragraph{Worlds}
But what does a ``bound on the authority'' of an executing block of code mean?
In our setting, there are no externally observable side-effects and the only
primitive authority that code may hold is authority over memory. As such, the
authority bounds we consider are related to memory, but in a form that is more
fine-grained than standard read/write authority: a piece of code's authority can be bounded
by arbitrary memory invariants that it is required to respect. Specifically, we
will define worlds $W \in \Worlds$, which describe a set of memory invariants, and
our results will express authority bounds on code as \emph{safety with respect to such
a world,} i.e., the fact that the code respects the invariants registered in
the world.

\paragraph{Safe values}
So, let's say that we have a world $W$ expressing that the memory must contain
value $42$ at address $0$, may contain arbitrary values at
addresses $50$-$60$, a $\readwrite$ capability for address $0$ at address $73$, and
an integer at address $100$ that may only increase over time\footnote{Indeed, we
  will allow a notion of \emph{evolvable} invariants, aka \emph{protocols}, that can
  express such a temporal property.}. Our main theorem will state that if the
current register file only contains safe words (numbers or capabilities which
preserve the invariants in $W$ under any interaction), then an execution will
necessarily also preserve the memory invariants (irrespective of the instructions being
executed).

To make this more precise, we need to define the set $\stdvr(W) \in
\powerset{\Words}$ of words that are safe w.r.t.\ $W$. Essentially, the set
should only include words that preserve $W$'s invariants under any interaction,
but should otherwise be as liberal as possible. Numbers are clearly always safe,
as they cannot be used to break invariants. Whether a capability is safe depends
on the authority that it carries. In the above-described world, a read capability
for address $0$ is safe, as it can only be used to read the value $42$, which is
itself safe. However, a write capability for address $0$ is not safe: it can be
used to overwrite the memory at that address with a value other than $42$,
breaking the invariant for that address.

\paragraph{Step-indexing}
More generally, we want to define that a read capability for memory range
$[b,e]$ is safe, if the world guarantees that the words at those addresses are
themselves safe. However, this definition is cyclic: what if the world
guarantees that the memory at address $a$ will contain a read capability for
address $a$? The definition then just says that a read capability for address
$a$ is safe if and only if the same read capability for address $a$ is safe. This form of
cyclic reasoning is related to similar challenges in languages with recursive
types or higher-order ML-style references, and a standard solution is
to use step-indexing \cite{Appel:2001:IMR:504709.504712}: essentially, the cycle is broken by defining safety up to a
certain number of interaction steps. All words will be considered
safe up to $0$ steps (since if there is no interaction, nothing unsafe
can happen), and, for example, a read capability will be safe up to $n$
steps if the world guarantees that the words at the corresponding addresses are
safe up to $n - 1$ steps. We can then prove that the above read capability for
address $a$ is safe up to any number of steps.

\paragraph{Future worlds}
So worlds are defined as a set of invariants on the memory, but what if we
allocate fresh memory through malloc? We may want to establish new invariants
for this freshly allocated memory and be sure that the adversary will also
respect those (if we don't provide them with capabilities through which the new
invariants can be broken). To accommodate this, we allow worlds to evolve, for
example by adding additional invariants for freshly allocated memory. Formally,
we define valid ways for a world $W$ to evolve into a new world $W'$ through a
future-world relation $W' \future W$ and we ensure that the set of safe words in
world $W$ must remain safe in any future world $W'$. Defining safety w.r.t.\ a
notion of evolvable worlds makes our logical relation into a \emph{Kripke}
logical relation \cite{pitts_operational_1998}.

\paragraph{Invariants and Recursive Worlds}
So worlds group a set of memory invariants, but how are they actually
defined formally? We represent each invariant by a region
$r \in \Regions$. We will see later that regions contain state
machines to support a notion of evolvable invariants, but in every
state, they also contain a predicate $H$ that defines which memory
segments are acceptable in the current state of the invariant.
Unfortunately, it is not enough to just take
$H \in \powerset{\MemSegments}$, because sometimes the invariant may
itself be world-dependent. For example, we may want to express
invariants like ``the memory at address $50$ contains a value that is
safe in the current world''. As explained, worlds may evolve, and the
set of safe values may grow in future worlds, and therefore we need to
index $H$ over worlds, i.e., take
$H \in \Worlds \rightarrow \powerset{\MemSegments}$. We then end up
with worlds containing regions with world-indexed predicates,
i.e., the set of worlds must be recursively defined. We will see how such a recursive
definition can be accommodated using techniques from the
literature (essentially an advanced application of step-indexing).

\paragraph{Local capabilities}
A final thing we provide informal intuition for is how our results take into account local
capabilities and their special treatment by the hardware. Normally, when we
invoke an untrusted piece of code and provide it with certain global
capabilities, it may have stored those capabilities in memory and we will only
be able to reinvoke the code if we can guarantee that those values are still
valid. Formally, worlds represent the invariants that global capabilities'
safety relies on and the reinvocation is only safe in future worlds, where the
invariants are respected.

However, if we provide the adversary with local capabilities in that first
invocation, then the situation is a bit different. The adversary has no way to
store these local capabilities, so if we make sure that there are also no old
local capabilities in the register file for the second invocation (including the
capability being invoked), then the adversary cannot use them any more.%
\footnote{We are ignoring write-local capabilities in this discussion. If the
  adversary does have access to write-local capabilities in the first and second
  invocation, then the memory they address must entirely be cleared before the
  second invocation in order for the reinvocation to remain safe.} Therefore, we
can allow the second invocation to happen in \emph{private} future worlds ($W'
\futurestr W$), in which safe global capabilities remain safe, but local ones
don't. This private future world relation is more liberal than the standard,
public one ($W' \futurewk W$, in which \emph{all} safe capabilities remain
safe). Concretely, worlds may contain \emph{temporary} regions, representing
invariants that only local capabilities may rely on for their safety, and which
may be revoked (disabled) in private future worlds.

Interestingly, this idea is a variant of a notion of public/private future
worlds that has been previously used in the literature (see
\sectionname~\ref{sec:discussion}). However, temporary regions are new in our setting
and there is an interesting interplay with the recursiveness of the worlds: for
a temporary region, the predicate $H \in \Worlds \rightarrow
\powerset{\MemSegments}$ (which defines the safe memory segments in the
current world) is only required to be monotone w.r.t.\ public
future worlds (i.e.\ safe memory segments remain safe in public future worlds),
but for permanent regions, it must be monotone w.r.t.\ private future worlds.
As a consequence, the memory for a permanent region may not contain local
capabilities (as their safety would be broken in private future worlds), which
in turn implies that only local capabilities may have write-local permission (a
general sanity requirement when using local capabilities\footnote{In fact, local
  capabilities become useless as soon as the adversary has access to a single
  global, write-local capability.})\footnote{For explanation purposes, this
  discussion ignores certain ways to allow for local capabilities in a permanent
  region, for example, by not requiring that they are valid or requiring that
  they are local versions of valid global capabilities.}.

@@

As describe above, worlds are a collection of invariants. Formally, a world is a
finite map from region names, modeled as natural numbers, to regions that each
correspond to an invariant of part of the memory.  We have three types of
regions: \emph{permanent}, \emph{temporary}, and \emph{revoked}.
%$\perma$, $\temp$, and $\revoked$. The
%$\perma$ and $\temp$ 
Each permanent and temporary region contains a state transition system, with
public and private transitions, to describe how the invariants are allowed to
change over time. In other words, they are protocols for the region's
memory. These are similar to what has been used in logical relations for
high-level
languages~\cite{Ahmed:popl09,Dreyer:jfp12,Devriese:2016ObjCap}. Protocols
imposed by permanent regions stay in place indefinitely. Any capability, local
or global, can depend on these protocols. Protocols imposed by temporary regions
can be revoked in private future worlds. Doing this may break the safety of
local capabilities but not global ones. This means that local capabilities
can safely depend on the protocols imposed by temporary regions, but global
capabilities cannot, since a global capability may outlive a temporary region
that is revoked. This is illustrated in \figurename~\ref{fig:cap-world}.

@@

%\dominique{``end up being defined recursively'' does not sound very insightful}
% A region defines what its part of memory can legally look like. However,
% validity of the memory's contents may itself depend on the state of other parts
% of memory. To accommodate this, worlds need to be defined recursively: Worlds
% describe memory, which can contain capabilities, whose safety is
% world-dependent. 
As described in \sectionname~\ref{sec:formalizing-guarantees}, regions define sets of
safe memory segments, but this set may itself be world-dependent. In other
words, our worlds are defined recursively. Recursive worlds are a common
requirement in Kripke models and the following lemma uses the method of
\cite{Birkedal:2011:SKM:1926385.1926401,Birkedal:tutorial-notes} for
constructing them. The formulation of the lemma is technical, so we recommend
that non-expert readers ignore the technicalities and accept that there exists a
set of worlds $\Wor$ and two relations $\futurestr$ and $\futurewk$ satisfying
the (recursive) equations in the theorem (where the $\blater$ operator can be
safely ignored).
% Short technical explanation
% The recursiveness of the worlds gives rise to a recurisve domain
% equation. Theorem~\ref{thm:world-existence} gives the solution to this
% recursive domain equation.
% \dominique{it is not explained what cofes are?}
% \lau{I added what cofe means but no formal definition (I think this goes under the category technicalities)}
\begin{theorem}\label{thm:world-existence}
  There exists a \cofe{} (complete ordered family of equivalences) $\Wor$ and preorders $\futurestr$ and
  $\futurewk$ such that $(\Wor,\futurestr)$ and $(\Wor,\futurewk)$ are
  preordered \cofes{}, and there exists an isomorphism $\xi$ such that
  \begin{align*}
      \xi \; : \; & \Wor \cong \blater (\nats \finparfun \Regions)\\
      \Regions  = & \; \{\revoked\} \uplus \\
               & \; \{\temp\} \times \States \times \Rels \times (\States \fun (\Wor \monwknefun \UPred{\HeapSegments}))\uplus \\
               & \; \{\perma\} \times \States \times \Rels \times (\States \fun (\Wor \monstrnefun \UPred{\HeapSegments}))
    \end{align*}
  and for $W, W' \in \Wor\ldotp \quad  
  \begin{aligned}
    W' \futurestr W & \Leftrightarrow \xi(W') \futurestr \xi(W)   \\
    W' \futurewk W & \Leftrightarrow \xi(W') \futurewk \xi(W)
  \end{aligned}$
\end{theorem}
In the above theorem, $\States \times \Rels$ corresponds to the aforementioned
state transition system where $\Rels$ contains pairs of relations corresponding
to the public and private transitions, and $\States$ is an unspecified set that
we assume to contain at least the states we use in this paper. The last part of
the temporary and permanent regions is a state interpretation function that
determines what memory segments the region permits in each state of the state
transition system.  The different monotonicity requirements in the two
interpretation functions reflects how permanent regions rely only on permanent
protocols whereas temporary regions can rely on both temporary and permanent
protocols.  $\UPred{\HeapSegments}$ is a step-indexed, downwards closed
predicate on memory segments.

@@

\subsubsection{World satisfaction}
A memory satisfies a world, written $\memSat{\ms}{W}$, if it can be partitioned
into disjoint parts such that each part is accepted by an active (permanent or
temporary) region.  Revoked regions are not taken into account as their memory protocols are no longer in effect.
\begin{align*}
  &\memSat{\ms}{W}
    \text{ iff }\quad \left\{\begin{aligned}
        &\exists P : \activeReg{W} \rightarrow \HeapSegments \ldotp \hs = \biguplus_{r\in\activeReg{W}}P(r) \text{ and } \\[-.35cm]
        &\quad \forall r \in \activeReg{W} \ldotp\\
        &\quad \quad \exists H,s \ldotp W(r) = (\_,s,\_,\_,H) \text{ and } \npair[n]{P(r)} \in H(s)(\xi^{-1}(W))\\
      \end{aligned}\right.
\end{align*}
Note that it is possible to state nonsensical worlds with conflicting
requirements for the same part of memory. There would, however, be no
memory that would satisfy such a world which for all intents and purposes would
make it discounted by the logical relation.

@@

\begin{figure}[htbp]
  \centering
  \begin{align*}
   \readCond{}(\gl)(W) &=  \left\{ \npair{(\start,\addrend)} \; \middle| \;
    \begin{array}{l}
       \exists r \in \var{localityReg}(g,W) \ldotp \\
\quad   \exists [\start',\addrend'] \supseteq [\start,\addrend] \ldotp W(r)\nsubsim[n] \iota_{\start',\addrend'}^\pwl 
    \end{array}
    \right\}\\
   \writeCond{}(\iota,\gl)(W) &=  \left\{
    \npair{(\start,\addrend)}
    \; \middle| \;
    \begin{aligned}
      & \exists r \in \var{localityReg}(g,W) \ldotp \\
      & W(r) \text{ is address-stratified and} \\[-0.2cm]
      & \;\; \exists [\start',\addrend'] \supseteq [\start,\addrend] \ldotp W(r)\nsupsim[n-1] \iota_{\start',\addrend'}
    \end{aligned} \right\}\\
   \execCond{}(\gl)(W) &= 
    \left\{
      \npairP{(\var{P}, \start,\addrend)}
     \; \middle| \;
    \begin{aligned}
      & \forall n' < n, W' \future W, a \in [\start,\addrend], \perm \in \var{P} \ldotp \\
      & \;\;\;\; \npair[n']{((\perm,\gl),\start,\addrend,\addr)} \in \stder(W')\\
    \end{aligned} \right\} \\
   \entryCond{}(\gl)(W) &= 
    \left\{ \npair{(\start,\addrend,\addr)} \; \middle| \;
    \begin{array}{l}
      \forall n' < n \ldotp \forall W' \future W \ldotp\\
      \quad \npair[n']{((\exec,\gl),\start,\addrend,\addr)} \in \stder(W')
    \end{array}
    \right\} \\
   \text{where } \gl = \local \Rightarrow \future = \futurewk \text{ and } \gl = \glob \Rightarrow \future = \futurestr \span
  \end{align*}
\caption{Permission-based conditions}
\label{fig:perm-conds}
\end{figure}
}
%subsection introduction
% \dominique{I would try to motivate the LR defs by saying that the LR is
%   constructed to express the FTLR and the FTLR says that if a piece of code only
%   has access to values respecting the invariants in a world, then it will
%   respect them itself too.} 
The logical relation defines semantically when values, program counters, and
configurations are safe according to the intuitions described in
\sectionname~\ref{sec:formalizing-guarantees}. The definition is found in
\figurename{}s~\ref{fig:logrel} and~\ref{fig:perm-conds} and we provide some
explanations in the following paragraphs. We omit some definitions and explain
them only verbally, but precise definitions can be found in the technical
appendix~\cite{technical_appendix}.

First, the \emph{observation relation} $\observations$ defines what
configurations we consider safe. As previously stated, a configuration is safe
with respect to a world when the execution of said configuration does not break
the memory protocols of the world. Roughly speaking, this means that when the
execution of a configuration halts, then there is a private future world that
the resulting memory satisfies. Notice that failing is considered safe behavior.
In fact, the machine often resorts to failing when an unauthorized access is
attempted, like loading from a capability without read permission. This is
similar to \cite{Devriese:2016ObjCap}'s logical relation for an untyped
language, but unlike typical logical relations for typed languages, which
require that programs do not fail.

@@

The \emph{value relation} $\stdvr$ defines when words are safe, i.e.\ when they
cannot be used to break the memory protocols defined by the world. We make the
value relation as liberal as possible by considering what is the most we can
allow an adversary to use a capability for without breaking the memory
protocols. Non-capability data is always safe because it provides no
authority. Capabilities give the authority to manipulate memory and potentially
break memory protocols, so they need to satisfy certain conditions to be
safe. In \figurename~\ref{fig:perm-conds}, we define such a condition for each
kind of permission a capability can have.

>>>>

\section{Logical relation}
\label{sec:logical-relation}

In this section, we formalize the guarantees provided by the
capability machine, including the specific guarantees for local
capabilities, by means of a step-indexed Kripke logical relation 
with recursively defined worlds. 
We use the logical relation in the following section to show
local-state encapsulation and control-flow integrity properties for
challenging example programs.

% Our logical relation will prescribe which values are capability safe given a
% world $W$ containing a set of memory invariants. For expressiveness,
% we use invariants which can evolve over time (sometimes also referred
% to as \emph{protocols}). We thus start by describing the worlds we
% will use.

% Now that we have defined our calling convention, how can we be sure that it
% works? More concretely, suppose that we have a program that uses the convention
% in its interaction with untrusted adversary code. Can we formally prove the
% program's correctness if it relies on well-bracketed control flow and private
% state encapsulation for the interaction with the adversary? Clearly, such a
% proof should depend on a formal expression of the guarantees provided by the
% capability machine, including the specific guarantees for local capabilities.

% In this section, we construct such a formalization. We make use of some
% well-studied and powerful (but non-trivial) machinery from the literature.
% Specifically, we employ a unary step-indexed Kripke logical relation with
% recursive worlds, and some additional special characteristics of our own.
% Step-indexing, Kripke logical relations and recursive worlds are techniques that
% may be familiar from lambda calculus settings, but it may not be clear to the
% reader how they apply in this more low-level assembly language. Therefore, we do
% not immediately dive into the details, but we first try to provide some informal
% intuition about how all of this machinery comes into play in our setting, in the
% next section.

% Note: even though the calling convention is the main application in this paper,
% the logical relation we construct is very general and should be regarded as an
% independent contribution.

% \subsection{Formalizing the guarantees of the capability machine}
% \label{sec:formalizing-guarantees}
% What differentiates a capability machine from a more standard assembly language
% is that we can bound the authority of an executing block of code, based solely
% on the capabilities it has access to. Specifically, it does not matter which
% instructions are actually executed, i.e., the bound also applies to untrusted
% adversary code that has not been inspected or modified in any way.

% \paragraph{Worlds}
% But what does a ``bound on the authority'' of an executing block of code mean?
% In our setting, there are no externally observable side-effects and the only
% primitive authority that code may hold is authority over memory. As such, the
% authority bounds we consider are related to memory, but in a form that is more
% fine-grained than standard read/write authority: a piece of code's authority can be bounded
% by arbitrary memory invariants that it is required to respect. Specifically, we
% will define worlds $W \in \Worlds$, which describe a set of memory invariants, and
% our results will express authority bounds on code as \emph{safety with respect to such
% a world,} i.e., the fact that the code respects the invariants registered in
% the world.

% \paragraph{Safe values}
% So, let's say that we have a world $W$ expressing that the memory must contain
% value $42$ at address $0$, may contain arbitrary values at
% addresses $50$-$60$, a $\readwrite$ capability for address $0$ at address $73$, and
% an integer at address $100$ that may only increase over time\footnote{Indeed, we
%   will allow a notion of \emph{evolvable} invariants, aka \emph{protocols}, that can
%   express such a temporal property.}. Our main theorem will state that if the
% current register file only contains safe words (numbers or capabilities which
% preserve the invariants in $W$ under any interaction), then an execution will
% necessarily also preserve the memory invariants (irrespective of the instructions being
% executed).

% To make this more precise, we need to define the set $\stdvr(W) \in
% \powerset{\Words}$ of words that are safe w.r.t.\ $W$. Essentially, the set
% should only include words that preserve $W$'s invariants under any interaction,
% but should otherwise be as liberal as possible. Numbers are clearly always safe,
% as they cannot be used to break invariants. Whether a capability is safe depends
% on the authority that it carries. In the above-described world, a read capability
% for address $0$ is safe, as it can only be used to read the value $42$, which is
% itself safe. However, a write capability for address $0$ is not safe: it can be
% used to overwrite the memory at that address with a value other than $42$,
% breaking the invariant for that address.

% \paragraph{Step-indexing}
% More generally, we want to define that a read capability for memory range
% $[b,e]$ is safe, if the world guarantees that the words at those addresses are
% themselves safe. However, this definition is cyclic: what if the world
% guarantees that the memory at address $a$ will contain a read capability for
% address $a$? The definition then just says that a read capability for address
% $a$ is safe if and only if the same read capability for address $a$ is safe. This form of
% cyclic reasoning is related to similar challenges in languages with recursive
% types or higher-order ML-style references, and a standard solution is
% to use step-indexing \cite{Appel:2001:IMR:504709.504712}: essentially, the cycle is broken by defining safety up to a
% certain number of interaction steps. All words will be considered
% safe up to $0$ steps (since if there is no interaction, nothing unsafe
% can happen), and, for example, a read capability will be safe up to $n$
% steps if the world guarantees that the words at the corresponding addresses are
% safe up to $n - 1$ steps. We can then prove that the above read capability for
% address $a$ is safe up to any number of steps.

% \paragraph{Future worlds}
% So worlds are defined as a set of invariants on the memory, but what if we
% allocate fresh memory through malloc? We may want to establish new invariants
% for this freshly allocated memory and be sure that the adversary will also
% respect those (if we don't provide them with capabilities through which the new
% invariants can be broken). To accommodate this, we allow worlds to evolve, for
% example by adding additional invariants for freshly allocated memory. Formally,
% we define valid ways for a world $W$ to evolve into a new world $W'$ through a
% future-world relation $W' \future W$ and we ensure that the set of safe words in
% world $W$ must remain safe in any future world $W'$. Defining safety w.r.t.\ a
% notion of evolvable worlds makes our logical relation into a \emph{Kripke}
% logical relation \cite{pitts_operational_1998}.

% \paragraph{Invariants and Recursive Worlds}
% So worlds group a set of memory invariants, but how are they actually
% defined formally? We represent each invariant by a region
% $r \in \Regions$. We will see later that regions contain state
% machines to support a notion of evolvable invariants, but in every
% state, they also contain a predicate $H$ that defines which memory
% segments are acceptable in the current state of the invariant.
% Unfortunately, it is not enough to just take
% $H \in \powerset{\MemSegments}$, because sometimes the invariant may
% itself be world-dependent. For example, we may want to express
% invariants like ``the memory at address $50$ contains a value that is
% safe in the current world''. As explained, worlds may evolve, and the
% set of safe values may grow in future worlds, and therefore we need to
% index $H$ over worlds, i.e., take
% $H \in \Worlds \rightarrow \powerset{\MemSegments}$. We then end up
% with worlds containing regions with world-indexed predicates,
% i.e., the set of worlds must be recursively defined. We will see how such a recursive
% definition can be accommodated using techniques from the
% literature (essentially an advanced application of step-indexing).

% \paragraph{Local capabilities}
% A final thing we provide informal intuition for is how our results take into account local
% capabilities and their special treatment by the hardware. Normally, when we
% invoke an untrusted piece of code and provide it with certain global
% capabilities, it may have stored those capabilities in memory and we will only
% be able to reinvoke the code if we can guarantee that those values are still
% valid. Formally, worlds represent the invariants that global capabilities'
% safety relies on and the reinvocation is only safe in future worlds, where the
% invariants are respected.

% However, if we provide the adversary with local capabilities in that first
% invocation, then the situation is a bit different. The adversary has no way to
% store these local capabilities, so if we make sure that there are also no old
% local capabilities in the register file for the second invocation (including the
% capability being invoked), then the adversary cannot use them any more.%
% \footnote{We are ignoring write-local capabilities in this discussion. If the
%   adversary does have access to write-local capabilities in the first and second
%   invocation, then the memory they address must entirely be cleared before the
%   second invocation in order for the reinvocation to remain safe.} Therefore, we
% can allow the second invocation to happen in \emph{private} future worlds ($W'
% \futurestr W$), in which safe global capabilities remain safe, but local ones
% don't. This private future world relation is more liberal than the standard,
% public one ($W' \futurewk W$, in which \emph{all} safe capabilities remain
% safe). Concretely, worlds may contain \emph{temporary} regions, representing
% invariants that only local capabilities may rely on for their safety, and which
% may be revoked (disabled) in private future worlds.

% Interestingly, this idea is a variant of a notion of public/private future
% worlds that has been previously used in the literature (see
% \sectionname~\ref{sec:discussion}). However, temporary regions are new in our setting
% and there is an interesting interplay with the recursiveness of the worlds: for
% a temporary region, the predicate $H \in \Worlds \rightarrow
% \powerset{\MemSegments}$ (which defines the safe memory segments in the
% current world) is only required to be monotone w.r.t.\ public
% future worlds (i.e.\ safe memory segments remain safe in public future worlds),
% but for permanent regions, it must be monotone w.r.t.\ private future worlds.
% As a consequence, the memory for a permanent region may not contain local
% capabilities (as their safety would be broken in private future worlds), which
% in turn implies that only local capabilities may have write-local permission (a
% general sanity requirement when using local capabilities\footnote{In fact, local
%   capabilities become useless as soon as the adversary has access to a single
%   global, write-local capability.})\footnote{For explanation purposes, this
%   discussion ignores certain ways to allow for local capabilities in a permanent
%   region, for example, by not requiring that they are valid or requiring that
%   they are local versions of valid global capabilities.}.

@@

%As describe above, worlds are a collection of invariants. Formally,

A world is a
finite map from region names, modeled as natural numbers, to regions that each
correspond to an invariant of part of the memory.  We have three types of
regions: \emph{permanent}, \emph{temporary}, and \emph{revoked}.
%$\perma$, $\temp$, and $\revoked$. The
%$\perma$ and $\temp$ 
Each permanent and temporary region contains a state transition system, with
public and private transitions, to describe how the invariants are allowed to
change over time. In other words, they are protocols for the region's
memory. These are similar to what has been used in logical relations for
high-level
languages~\cite{Ahmed:popl09,Dreyer:jfp12,Devriese:2016ObjCap}. Protocols
imposed by permanent regions stay in place indefinitely. Any capability, local
or global, can depend on these protocols. Protocols imposed by temporary regions
can be revoked in private future worlds. Doing this may break the safety of
local capabilities but not global ones. This means that local capabilities
can safely depend on the protocols imposed by temporary regions, but global
capabilities cannot, since a global capability may outlive a temporary region
that is revoked. This is illustrated in \figurename~\ref{fig:cap-world}.

@@

%\dominique{``end up being defined recursively'' does not sound very insightful}
% A region defines what its part of memory can legally look like. However,
% validity of the memory's contents may itself depend on the state of other parts
% of memory. To accommodate this, worlds need to be defined recursively: Worlds
% describe memory, which can contain capabilities, whose safety is
% world-dependent. 
%As described in \sectionname~\ref{sec:formalizing-guarantees}, regions define sets of

Regions are used to define safe memory segments, but this set may itself be world-dependent. In other
words, our worlds are defined recursively. Recursive worlds are common
in Kripke models and the following lemma uses the method of
\cite{Birkedal:2011:SKM:1926385.1926401,Birkedal:tutorial-notes} for
constructing them. The formulation of the lemma is technical, so we recommend
that non-expert readers ignore the technicalities and accept that there exists a
set of worlds $\Wor$ and two relations $\futurestr$ and $\futurewk$ satisfying
the (recursive) equations in the theorem (where the $\blater$ operator can be
safely ignored).
% Short technical explanation
% The recursiveness of the worlds gives rise to a recurisve domain
% equation. Theorem~\ref{thm:world-existence} gives the solution to this
% recursive domain equation.
% \dominique{it is not explained what cofes are?}
% \lau{I added what cofe means but no formal definition (I think this goes under the category technicalities)}
\begin{theorem}\label{thm:world-existence}
  There exists a \cofe{} (complete ordered family of equivalences) $\Wor$ and preorders $\futurestr$ and
  $\futurewk$ such that $(\Wor,\futurestr)$ and $(\Wor,\futurewk)$ are
  preordered \cofes{}, and there exists an isomorphism $\xi$ such that
  \begin{align*}
      \xi \; : \; & \Wor \cong \blater (\nats \finparfun \Regions)\\
      \Regions  = & \; \{\revoked\} \uplus \\
               & \; \{\temp\} \times \States \times \Rels \times (\States \fun (\Wor \monwknefun \UPred{\HeapSegments}))\uplus \\
               & \; \{\perma\} \times \States \times \Rels \times (\States \fun (\Wor \monstrnefun \UPred{\HeapSegments}))
    \end{align*}
  and for $W, W' \in \Wor\ldotp \quad  
  \begin{aligned}
    W' \futurestr W & \Leftrightarrow \xi(W') \futurestr \xi(W)   \\
    W' \futurewk W & \Leftrightarrow \xi(W') \futurewk \xi(W)
  \end{aligned}$
\end{theorem}
In the above theorem, $\States \times \Rels$ corresponds to the aforementioned
state transition system where $\Rels$ contains pairs of relations corresponding
to the public and private transitions, and $\States$ is an unspecified set that
we assume to contain at least the states we use in this paper. The last part of
the temporary and permanent regions is a state interpretation function that
determines what memory segments the region permits in each state of the state
transition system.  The different monotonicity requirements in the two
interpretation functions reflects how permanent regions rely only on permanent
protocols whereas temporary regions can rely on both temporary and permanent
protocols.  $\UPred{\HeapSegments}$ is the set of step-indexed, downwards closed
predicates on memory segments:
$\UPred{\HeapSegments} = \{ A \subseteq  \nats \times
\HeapSegments \mid 
\forall (n, ms) \in A. \forall m
\leq n. (m, ms) \in A\}$. 
\

@@

\subsubsection{World satisfaction}
A memory satisfies a world, written $\memSat{\ms}{W}$, if it can be partitioned
into disjoint parts such that each part is accepted by an active (permanent or
temporary) region.  Revoked regions are not taken into account as their memory protocols are no longer in effect.
\begin{align*}
  &\memSat{\ms}{W}
    \text{ iff }\quad \left\{\begin{aligned}
        &\exists P : \activeReg{W} \rightarrow \HeapSegments \ldotp \hs = \biguplus_{r\in\activeReg{W}}P(r) \text{ and } \\[-.35cm]
        &\quad \forall r \in \activeReg{W} \ldotp\\
        &\quad \quad \exists H,s \ldotp W(r) = (\_,s,\_,\_,H) \text{ and } \npair[n]{P(r)} \in H(s)(\xi^{-1}(W))\\
      \end{aligned}\right.
\end{align*}
% LB: omitted the following for brevity:
% Note that it is possible to state nonsensical worlds with conflicting
% requirements for the same part of memory. There would, however, be no
% memory that would satisfy such a world which for all intents and purposes would
% make it discounted by the logical relation.

@@

\begin{figure}[htbp]
  \centering
  \begin{align*}
   \readCond{}(\gl)(W) &=  \left\{ \npair{(\start,\addrend)} \; \middle| \;
    \begin{array}{l}
       \exists r \in \var{localityReg}(g,W) \ldotp \\
\quad   \exists [\start',\addrend'] \supseteq [\start,\addrend] \ldotp W(r)\nsubsim[n] \iota_{\start',\addrend'}^\pwl 
    \end{array}
    \right\}\\
   \writeCond{}(\iota,\gl)(W) &=  \left\{
    \npair{(\start,\addrend)}
    \; \middle| \;
    \begin{aligned}
      & \exists r \in \var{localityReg}(g,W) \ldotp \\
      & W(r) \text{ is address-stratified and} \\[-0.2cm]
      & \;\; \exists [\start',\addrend'] \supseteq [\start,\addrend] \ldotp W(r)\nsupsim[n-1] \iota_{\start',\addrend'}
    \end{aligned} \right\}\\
   \execCond{}(\gl)(W) &= 
    \left\{
      \npairP{(\var{P}, \start,\addrend)}
     \; \middle| \;
    \begin{aligned}
      & \forall n' < n, W' \future W, a \in [\start,\addrend], \perm \in \var{P} \ldotp \\
      & \;\;\;\; \npair[n']{((\perm,\gl),\start,\addrend,\addr)} \in \stder(W')\\
    \end{aligned} \right\} \\
   \entryCond{}(\gl)(W) &= 
    \left\{ \npair{(\start,\addrend,\addr)} \; \middle| \;
    \begin{array}{l}
      \forall n' < n \ldotp \forall W' \future W \ldotp\\
      \quad \npair[n']{((\exec,\gl),\start,\addrend,\addr)} \in \stder(W')
    \end{array}
    \right\} \\
   \text{where } \gl = \local \Rightarrow \future = \futurewk \text{ and } \gl = \glob \Rightarrow \future = \futurestr \span
  \end{align*}
\caption{Permission-based conditions}
\label{fig:perm-conds}
\end{figure}
}
%subsection introduction
% \dominique{I would try to motivate the LR defs by saying that the LR is
%   constructed to express the FTLR and the FTLR says that if a piece of code only
%   has access to values respecting the invariants in a world, then it will
%   respect them itself too.} 
The logical relation defines semantically when values, program
counters, and configurations are capability safe. The definition is
found in \figurename{}s~\ref{fig:logrel} and~\ref{fig:perm-conds} and
we provide some explanations in the following paragraphs. For space
reasons, we omit some definitions and explain them only verbally, but
precise definitions can be found in the technical
appendix~\cite{technical_appendix}.

First, the \emph{observation relation} $\observations$ defines what
configurations we consider safe. A configuration is safe
with respect to a world, when the execution of said configuration does not break
the memory protocols of the world. Roughly speaking, this means that when the
execution of a configuration halts, then there is a private future world that
the resulting memory satisfies. Notice that failing is considered safe behavior.
In fact, the machine often resorts to failing when an unauthorized access is
attempted, like loading from a capability without read permission. This is
similar to \cite{Devriese:2016ObjCap}'s logical relation for an untyped
language, but unlike typical logical relations for typed languages, which
require that programs do not fail.

@@

The \emph{value relation} $\stdvr$ defines when words are safe. We make the
value relation as liberal as possible by considering what is the most we can
allow an adversary to use a capability for without breaking the memory
protocols. Non-capability data is always safe because it provides no
authority. Capabilities give the authority to manipulate memory and potentially
break memory protocols, so they need to satisfy certain conditions to be
safe. In \figurename~\ref{fig:perm-conds}, we define such a condition for each
kind of permission a capability can have.
{
For capabilities with read permission, the $\readCond{}$ ensures that it can
only be used to read safe words, i.e.\ words in the value relation. To guarantee
this, we require that the addressed memory is governed by a region $W(r)$ that
imposes safety as a requirement on the values contained. This safety requirement
is formulated in terms of a standard region $\iota^\pwl_{\start,\addrend}$. The
definition of that standard region is omitted for space reasons, but it simply
requires all the words in the range $[\start,\addrend]$ to be safe, i.e.\ in the
value relation. Requiring that $W(r)\nsubsim[n]\iota^\pwl_{\start,\addrend}$
means that $W(r)$ must accept only safe values like
$\iota^\pwl_{\start,\addrend}$, but can be even more restrictive if desired. The
read condition also takes into account the locality of the capability because,
generally speaking, global capabilities should only depend on permanent regions.
Concretely, we use the function $\var{localityReg}(\gl,W)$, which projects out
all active (non-revoked) regions when the locality $\gl$ is local, but only the
permanent regions when $\gl$ is global. The 
the definition of the standard region
$\iota^\pwl_{\start,\addrend}$ can be found
in~\cite{technical_appendix}; it makes use of the isomorphism from
Theorem~\ref{thm: world-existence}.
